### 2021-06-10 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Added index of notation. Uses makeindex.

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x, y, z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why this is a practical way to define our domain.

It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this thesis we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $\rho(z) := \sqrt{1-z^2}$.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere. These examples consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): written in spherical coordinates this is
\bseqnnumber{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo \label{eqn:laplacebeltrami}
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for some function $f$ where $\xvec := (x,y,z) \in \R^3$.

@@

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[table 18.3.1, eqn 18.9.15]{DLMF}:
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the $l$ degree \textit{Jacobi polynomial}, and $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)} \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'}.
}
Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are solutions to the differential equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0
}
which is helpfully known as the \textit{Legendre equation} \cite[14.2.1]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p5]{belousov1962tables}:
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

However we reiterate our goal here is to use techniques that can generalise to spectral methods on other domains using multidimensional OPs, and thus we proceed by exploiting the three-term recurrences and other relations of the univariate OPs that make up the spherical harmonics as we have defined them. With that being said, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$ the following identities hold:
\bseqn{
	&1)\quad P_{n}^{(\alpha, \beta)}(z) = {1 \over 2n + \alpha + \beta + 1} \Big\{- {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  					& \quadfour \quadtwo \quad + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									 					& \quadfour \quadtwo \quad + {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\ \\
	&2)\quad (1-z^2) \, P_{n}^{(\alpha, \beta)}(z) = {4 \over 2n + \alpha + \beta + 1} \Big\{{(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\ \\
	&3)\quad z \, P_{n}^{(\alpha, \beta)}(z) = {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
								& \quadeight + {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
								& \quadeight + {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
}
\end{lemma}
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a helpful corollary specific for our needs.
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
\bseqn{
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
where
\bseqn{
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1}, \quad
	&&\tilde \alpha_{l,m,2} := 
		\begin{cases}
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
		\end{cases}, \\
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)}, \quad
	&&\tilde \alpha_{l,m,4} := {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)}, \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases}, \quad
	&&\tilde \gamma_{l,m,2} := \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
}
\end{corollary}
\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x,y,z$ of the OPs. Define for $l \in \No$:
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
Note that these OP vectors we have defined are grouped and ordered according to the degree $l$, and within each degree ordered by the Fourier mode $m$. Let the Jacobi matrices $J_x, J_y, J_z$ hence be given by 
\bseqnnumber{
	J_x \SHvec(x,y,z) = x \SHvec(x,y,z), \quad J_y \SHvec(x,y,z) = y \SHvec(x,y,z), \quad J_z \SHvec(x,y,z) = z \SHvec(x,y,z) \label{eqn:jacobimatsdef}
}
for $(x,y,z) \in \Omega$. These Jacobi matrices can act on the coefficients vector of a function's expansion in the spherical harmonic basis to implement the operation for multiplication by $x$, $y$ or $z$. For example, given sufficiently large $n \in \N$ let the function $f(x,y,z) : \Omega \to \C$ be given by its expansion
\bseqn{
	f(x,y,z) = \sum_{l=0}^{N} \sum_{m=-l}^{l} f_{l,m} \; \Ylm(x,y,z) = \SHvec(x,y,z)^\top \: \fvec
}
where 
\bseqn{
	\fvec_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			\vdots \\
			f_{l,-l}
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\fvec := 
		\begin{pmatrix}
			\fvec_0 \\
			\vdots \\
			\fvec_N
		\end{pmatrix}.
}
We say that $\fvec$ is the \textit{coefficients vector} of the function $f$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J_x^\top \: \fvec$. In other words, $J_x^\top \: \fvec$ is the coefficients vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

	For $l \in \N$ and $m \in \Z$ s.t. $-l \le m \le l$, the vector spherical harmonics satisfy the relationships:

@@

As for the scalar case, these coefficients detailed above are then the nonzero entries of the Jacobi matrices for the VSHs. Define $\VSHvecfull$ by
\bseqnnumber{
	\VSHvecfull = 
		\begin{pmatrix} 
			\VSHvecfull_0 \\ \VSHvecfull_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecfulll = 
		\begin{pmatrix} 
			(\gradY^{-l}_l)^\top \\ (\gradpY^{-l}_l)^\top \\ \vdots \\ (\gradY^{l}_l)^\top \\ (\gradpY^{l}_l)^\top
		\end{pmatrix} \in \C^{2(2l+1) \times 3} \quad \forall \: l \in \No. \label{eqn:VSHfulldef}
}

>>>>

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x, y, z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain \nomenclature[Omega]{$\Omega$}{The domain of interest}
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why this is a practical way to define our domain.

It is useful to be able to transform between these cartesian coordinates and the spherical coordinates \nomenclature[phi]{$\varphi$}{The polar angle in spherical coordinates} \nomenclature[theta]{$\theta$}{The azimuthal angle in spherical coordinates}($\varphi, \theta$). Throughout this thesis we will use the convention that the spherical coordinate angles be defined by \nomenclature[xyz]{$x,y,z$}{Lower case non-bold $x,y,z$ represents cartesian coordinates}
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $\rho(z) := \sqrt{1-z^2}$.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere. These examples consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): written in spherical coordinates this is \nomenclature[Deltas]{$\DeltaS$}{The Laplace--Beltrami operator}
\bseqnnumber{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo \label{eqn:laplacebeltrami}
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for some function $f$ where $\xvec := (x,y,z) \in \R^3$.

@@

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[table 18.3.1, eqn 18.9.15]{DLMF}: \nomenclature[Pnab]{$P^{(a,b)}_n$}{The degree $n$ Jacobi polynomial (context determines normalized or not) unless otherwise stated}
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the $l$ degree \textit{Jacobi polynomial}, and \nomenclature[Pl]{$P_l$}{The degree $l$ Legendre polynomial} $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$ \nomenclature[Kronecker delta]{$\delta_{n, n'}$}{The Kronecker delta function; $1$ if $n=n'$, else $0$}
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)} \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'}.
}
Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are solutions to the differential equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0
}
which is helpfully known as the \textit{Legendre equation} \cite[14.2.1]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p5]{belousov1962tables}: \nomenclature[Plm]{$\Plm$}{The Associated Legendre polynomial of degree $l$ and order $m$}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}: \nomenclature[Ylm]{$\Ylm$}{The degree $l$ and order $m$ spherical harmonic}
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where \nomenclature[clm]{$\clm$}{The normalising constant for the spherical harmonic $\Ylm$}
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure \nomenclature[conjuagate]{$a^*$}{The complex conjugate of $a \in \C$}
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of \nomenclature[C]{$\C$}{The complex numbers} $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any \nomenclature[Z]{$\Z$}{The integers} $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in \nomenclature[R]{$\R$}{The real numbers} \nomenclature[Rd]{$\R^d$}{The d-dimensional coordinate space over the real numbers} $\R^3$.

@@

However we reiterate our goal here is to use techniques that can generalise to spectral methods on other domains using multidimensional OPs, and thus we proceed by exploiting the three-term recurrences and other relations of the univariate OPs that make up the spherical harmonics as we have defined them. With that being said, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$ the following identities hold:
	&1)\quad P_{n}^{(\alpha, \beta)}(z) = {1 \over 2n + \alpha + \beta + 1} \Big\{- {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  					& \quadfour \quadtwo \quad + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									 					& \quadfour \quadtwo \quad + {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\ \\
	&2)\quad (1-z^2) \, P_{n}^{(\alpha, \beta)}(z) = {4 \over 2n + \alpha + \beta + 1} \Big\{{(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\ \\
	&3)\quad z \, P_{n}^{(\alpha, \beta)}(z) = {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
								& \quadeight + {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
								& \quadeight + {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a helpful corollary specific for our needs. \nomenclature[alpha]{$\alpha_{l,m,j}$}{Lower case greek letters with subscripts refer to coefficients in scalar OP relations (unless otherwise stated)}
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1}, \quad
	&&\tilde \alpha_{l,m,2} := 
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)}, \quad
	&&\tilde \alpha_{l,m,4} := {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)}, \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases}, \quad
	&&\tilde \gamma_{l,m,2} := \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
\end{corollary}
\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x,y,z$ of the OPs. Define for $l \in \No$: \nomenclature[Pvector]{$\SHvec$}{Vector of the SH OPs} \nomenclature[Pvectorl]{$\SHvec_l$}{(Sub-)Vector of the degree $l$ SH OPs}
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
Note that these OP vectors we have defined are grouped and ordered according to the degree $l$, and within each degree ordered by the Fourier mode $m$. Let the Jacobi matrices \nomenclature[J]{$J_x, J_y, J_z$}{Jacobi operator matrices} $J_x, J_y, J_z$ hence be given by 
\bseqnnumber{
	J_x \SHvec(x,y,z) = x \SHvec(x,y,z), \quad J_y \SHvec(x,y,z) = y \SHvec(x,y,z), \quad J_z \SHvec(x,y,z) = z \SHvec(x,y,z) \label{eqn:jacobimatsdef}
}
for $(x,y,z) \in \Omega$. These Jacobi matrices can act on the coefficients vector of a function's expansion in the spherical harmonic basis to implement the operation for multiplication by $x$, $y$ or $z$. For example, given sufficiently large $n \in \N$ let the function $f(x,y,z) : \Omega \to \C$ be given by its expansion \nomenclature[fvec]{$\vec{f}, \vec{\xi}$}{Lower case bold letters refer to vectors} \nomenclature[transpose]{$\vec{v}^\top, A^\top$}{Transpose of the vector $\vec{v}$ or matrix $A$ (non complex conjuagate)}
\bseqn{
	f(x,y,z) = \sum_{l=0}^{N} \sum_{m=-l}^{l} f_{l,m} \; \Ylm(x,y,z) = \SHvec(x,y,z)^\top \: \fvec
}
where 
\bseqn{
	\fvec_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			\vdots \\
			f_{l,-l}
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\fvec := 
		\begin{pmatrix}
			\fvec_0 \\
			\vdots \\
			\fvec_N
		\end{pmatrix}.
}
We say that $\fvec$ is the \textit{coefficients vector} of the function $f$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J_x^\top \: \fvec$. In other words, $J_x^\top \: \fvec$ is the coefficients vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

	For $l \in \N$ and $m \in \Z$ s.t. $-l \le m \le l$, the vector spherical harmonics satisfy the relationships: \nomenclature[Alpha]{$A_{l,m,j}$}{Capital greek letters with subscripts refer to coefficients in vector OP (VSH) relations (unless otherwise stated)}

@@

As for the scalar case, these coefficients detailed above are then the nonzero entries of the Jacobi matrices for the VSHs. Define $\VSHvecfull$ by \nomenclature[T vector]{$\VSHvecfull$}{Vector of the VSHs} \nomenclature[Tl vector]{$\VSHvecfulll$}{(Sub)-Vector of the VSHs of order $l$} 
\bseqnnumber{
	\VSHvecfull = 
		\begin{pmatrix} 
			\VSHvecfull_0 \\ \VSHvecfull_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecfulll = 
		\begin{pmatrix} 
			(\gradY^{-l}_l)^\top \\ (\gradpY^{-l}_l)^\top \\ \vdots \\ (\gradY^{l}_l)^\top \\ (\gradpY^{l}_l)^\top
		\end{pmatrix} \in \C^{2(2l+1) \times 3} \quad \forall \: l \in \No. \label{eqn:VSHfulldef}
}
